---
title: "MINI-PROJECT 2: EDA and correlation analysis"
author: "Delphi Tooling Team"
date: today
execute:
  warning: false
  message: true
---

The `{InsightNetApr26}` package ensures all required Delphi tools are installed with the correct versions for this session.

```{r}
#| label: load-packages
if (!requireNamespace("pak", quietly = TRUE)) install.packages("pak")
if (!requireNamespace("InsightNetApr26", quietly = TRUE)) {
    pak::pkg_install("cmu-delphi/InsightNet-apr-2026/InsightNetApr26")
}
InsightNetApr26::verify_setup()

library(epidatr)
library(epiprocess)
library(dplyr)
library(ggplot2)
```

```{r}
#| label: setup
#| include: false
source(here::here("scripts", "setup.R"))
```

## 2.1 General Exploratory Data Analysis (EDA) utilities

⭐ Either continue with your data from Mini-Project 1 (in `epi_df` format) OR select a new data indicator of your choice and transform it to an `epi_df` using `epiprocess`. Make sure you pull the data for all states!

For example, we start with NHSN weekly RSV hospital admissions, pulling all states across multiple seasons.

> **Hint:** The epiprocess documentation and vignettes are available at [cmu-delphi.github.io/epiprocess](https://cmu-delphi.github.io/epiprocess).

```{r}
#| label: fetch-data
epidata_nhsn_state <- pub_covidcast(
  source = "nhsn",
  signal = "confirmed_admissions_rsv_ew",
  geo_type = "state",
  time_type = "week",
  geo_values = "ca,mi,ny,pa",
  time_values = epirange(202501, 202601)
) |>
  select(geo_value, time_value, value, issue)

edf <- epidata_nhsn_state |> 
  as_epi_df()
```

  - Check out your dataset! Type the object's name for a quick look. Then use `summary()` to view the dataset's metadata and overall statistics. For information on what each section means, visit <https://cmu-delphi.github.io/epiprocess/dev/reference/print.epi_df.html>

```{r}
#| label: eda-view
edf
summary(edf)
```

> 🛈 Look at the "Latency" section in the `summary()` output. The `(!)` symbol appears when there is a "notable" lag. For an up-to-date indicator, you would see no warning and typically a lag of only a few days. If your chosen indicator has a notable lag, how does that impact its use for real-time surveillance?

  - Use `epi_slide_mean()` and `epi_slide_sum()` to calculate moving averages and weekly totals in a new column of your `epi_df`. 
    - Try `.window_size` and `.align` parameters to control the sliding window size and the alignment of the sliding window, respectively.

> **Hint:** Because this data has `time_type = "week"`, `.window_size` must be expressed as a `difftime` (e.g., `as.difftime(3, units = "weeks")`).

```{r}
#| label: sliding
edf <- edf |>
  epi_slide_mean("value", .window_size = as.difftime(3, units = "weeks")) |>
   epi_slide_mean("value", .window_size = as.difftime(8, units = "weeks")) |>
  epi_slide_mean("value", .window_size = as.difftime(3, units = "weeks"), .align = "center",
    .new_col_names = "value_3wmean_center") |>
  epi_slide_sum("value", .window_size = as.difftime(4, units = "weeks"))

edf |>
  select(geo_value, time_value, value, value_3wav, value_8wav, value_3wmean_center, value_4wsum) |>
  head(10)
```

> 🛈 Plot the different columns you calculated, and compare the results. How does changing the `.window_size` from 1 to 8 weeks impact the "noise" in your plot? Is there a trade-off between smoothness and time-accuracy?

```{r}
#| label: sliding-plot
autoplot(edf, value, value_8wav)
```

  - Use `autoplot()` and `plot_heatmap()` to explore the spatial and temporal trends in the data.
  - `autoplot()`
    - Try `.facet_by = "geo_value"` to see each location in its own box.
    - Set `.interactive = TRUE` to enable zoom, hovering and data tooltips.
  - `plot_heatmap()`
    - Change the x-axis range using `+ coord_cartesian()`

```{r}
#| label: autoplot-heatmap
autoplot(edf, value, .facet_by = "geo_value")

autoplot(edf, value, .interactive = TRUE)

plot_heatmap(edf, value) +
  coord_cartesian(xlim = as.Date(c("2025-06-01", "2026-01-01")))
```

> 🛈 If not included already, try selecting all states and use `plot_heatmap()` to identify which states show the most synchronous highs or lows of the metric you're viewing. Can you spot any regional clusters or outliers? Do you see a pattern in where data is most frequently missing?

  - Estimate the indicator's relative change using `growth_rate()` and plot it side by side using `autoplot()`.

   - Filter your data to a state of interest (e.g. `"mi"`) and remove all `NA` values from the column `value`.
   - Arrange your data by `time_value` and then calculate the growth rate of `value` in a new column using `growth_rate()`.
    - Use `ggplot` to plot both the column `value` and your growth rate column together in one line plot — Notice anything odd?
    - Look at the scale of `value` vs. the scale of your growth rate column — make adjustments to your plot as necessary to view both values.

```{r}
#| label: growth-rate
edf |>
  filter(geo_value == "mi",
  !is.na(value)) |>
  mutate(gr = growth_rate(value)) |> 
  autoplot(value, gr)
```

> 🛈 What does the growth rate measure? How is it changing as your data changes?

⚔️ Explore `rtestim` to estimate the effective reproductive number, Rt, using an indicator based on actual case counts. The main entry point is `cv_estimate_rt()`, which selects the smoothness parameter through cross-validation. Pass it a numeric vector of case counts from a single location.

```{r}
#| label: rtestim
library(rtestim)

# This example uses the history of real Covid-19 case counts in Canada
# It is available with the rtestim package
mod_rt <- cv_estimate_rt(
  observed_counts = cancovid$incident_cases,
  x = cancovid$date
  )

plot(mod_rt, which_lambda = "lambda.1se")
```

## 2.2 Compare two different indicators

⭐ Explore other available indicators with the [Delphi EpiPortal](https://delphi.cmu.edu/epiportal/), or, follow along with these steps using the NCHS Influenza Deaths (Weekly new, per 100k people) and ILINet's weighted percent influenza-like illness rates for Pennsylvania.

  - Join these two datasets into a single wide format `epi_df`.

> **Hint:** Turn the two sets into individual `epi_df`s first, then join them [`epiprocess::as_epi_df()`].

```{r}
#| label: fetch-two-indicators
#| error: true

# NCHS Influenza Deaths: weekly new deaths per 100k
edf_flu_deaths <- pub_covidcast(
  source = "nchs-mortality",
  signal = "deaths_flu_incidence_prop",
  geo_type = "state",
  time_type = "week",
  geo_values = "ca,mi,ny,pa"
) |>
  select(geo_value, time_value, flu_deaths = value) |> 
  as_epi_df()

# ILINet: weighted percent influenza-like illness
edf_ili_raw <- pub_fluview(
  regions = "ca,mi,ny,pa",
  epiweeks = epirange(201001, 202520)
)

edf_ili <- edf_ili_raw |>
  select(geo_value = region, time_value = epiweek, wili) |>
  as_epi_df()
```

```{r}
#| label: join-indicators
#| error: true
edf_joined <- inner_join(
  edf_flu_deaths,
  edf_ili,
  by = c("geo_value", "time_value")
) |>
  drop_na()

edf_joined
```

  - Use `autoplot()` with both indicators selected as parameters to view the data on one plot.

```{r}
#| label: autoplot-two
#| error: true
autoplot(edf_joined, flu_deaths, wili)
```

  - Calculate a correlation using `epiprocess::epi_cor()`.

    - Set `cor_by` to `geo_value` and `time_value` to estimate the correlation over time within the same location (per-location temporal correlation) and across regions at a specific moment (correlation over time).
    - You can also vary the `method` (using `"pearson"` for linear relationship or `"spearman"` for ranked-based correlation).

> **Hint:** You can find the "correlation" vignettes [here](https://cmu-delphi.github.io/epiprocess/dev/articles/correlation.html).

```{r}
#| label: correlation
#| error: true
# Per-location temporal correlation
epi_cor(edf_joined, flu_deaths, wili,
  cor_by = geo_value
)

epi_cor(edf_joined, flu_deaths, wili,
  cor_by = time_value, method = "spearman"
)
epi_cor(edf_joined, flu_deaths, wili,
  cor_by = geo_value, method = "spearman"
)
```

Use `ggplot2` or any other plotting package to analyze how the correlation changes over time.

```{r}
#| label: correlation-plot
#| error: true
cor_spatial <- epi_cor(edf_joined, flu_deaths, wili,
  cor_by = time_value, method = "spearman"
) |> 
  drop_na()


ggplot(cor_spatial, aes(x = time_value, y = cor)) +
  geom_line(alpha = 0.5) +
  labs(x = NULL, y = "Spearman rho",
    title = "NCHS flu deaths vs. weighted ILI: spatial correlation over time")
```

⚔️ Use the `dt1` and `dt2` parameters in `epi_cor()` to find the exact lag that maximizes the correlation between your candidate and guiding indicators. You can even wrap this in a `lapply()` over a range of lags to build a "lag-lead" visualization of the relationship between the two signals.

```{r}
#| label: lagged-correlation
#| error: true

lag_summary <- lapply(-8:8, \(lag) {
  epi_cor(edf_joined, flu_deaths, wili,
    cor_by = geo_value, dt1 = -lag, method = "spearman"
  ) |>
    mutate(lag = lag)
}) |>
  bind_rows() |>
  group_by(lag) |>
  summarize(
    mean_cor = mean(cor, na.rm = TRUE),
    q25      = quantile(cor, 0.25, na.rm = TRUE),
    q75      = quantile(cor, 0.75, na.rm = TRUE),
    .groups  = "drop"
  )

ggplot(lag_summary, aes(x = lag)) +
  geom_line(aes(y = mean_cor)) +
  geom_point(aes(y = mean_cor)) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "Lag (weeks)", y = "Spearman correlation",
    title = "NCHS flu deaths vs. weighted ILI:\nmean correlation by lag")
```

> 🛈 If you see a high spatial correlation but a low temporal correlation, what does that tell you about the relationship between the two indicators? Is it possible for two signals to be highly correlated in space but not in time?

## 2.3 Revision behavior - Understanding how data matures over time

⭐ Fetch and visualize the version history of an indicator — you can choose your own indicator, or follow along with the Day-adjusted COVID-related Doctor Visits indicator.

> **Hint:** For version-aware processing, including revision analysis and modelling, the training data needs to include the revision history. Since we're using Delphi-hosted data right now, we can fetch data versions from the Delphi Epidata API by specifying `issues = "*"` in the API call.

```{r}
#| label: fetch-archive
archive_raw <- pub_covidcast(
  source = "doctor-visits",
  signal = "smoothed_adj_cli",
  geo_type = "state",
  time_type = "day",
  geo_values = c("mi", "ny", "tx", "pa"),
  time_values = epirange(20200101, 20200601),
  issues = "*"
)
```

  - Convert the data into `epi_archive` format using `as_epi_archive()`.

> **Hint:** `as_epi_archive()` requires a column called `version`. Since the Delphi Epidata API returns the `issue` column, you may need to rename this column with `rename(version = issue)`.

```{r}
#| label: to-archive
archive_data <- archive_raw |>
  select(geo_value, time_value, version = issue, value) |>
  as_epi_archive()
```

  - Plot the data using `autoplot()`.

> **Hint:** To view the help file for this method, type `?autoplot.epi_archive` in R or visit [this link](https://cmu-delphi.github.io/epiprocess/dev/reference/autoplot.epi_archive.html).

```{r}
#| label: autoplot-archive
autoplot(archive_data, value)
```

> 🛈 Look at the divergence in the `autoplot()` lines. Which version tends to be the most "optimistic" or "pessimistic" compared to the final data (in black)? Do you notice any systematic bias in the initial reports?

  - "Slice" the archive at a specific version date using `epix_as_of()`. [Documentation](https://cmu-delphi.github.io/epiprocess/dev/reference/epix_as_of.html)

```{r}
#| label: as-of
snapshot_early <- epix_as_of(archive_data, version = as.Date("2020-06-01"))
```

> 🛈 When you extract a snapshot using `epix_as_of()`, you are essentially "time-travelling" to see the state of knowledge on that date. If the preliminary values on a date were much lower than the final values, how might that lead to a false sense of security during an emerging surge?

> 🛈 How might this ability to view history and current in a report assist in decision-making around a particular metric? Do you handle this in any particular way at your institution/agency?

⚔️ Instead of fetching the entire history, use the `as_of` parameter in `pub_covidcast()` to fetch the data exactly as it was reported on a specific date. Convert it to an `epi_df` and compare it to the slice you took above.

```{r}
#| label: as-of-covidcast
edf_as_of <- pub_covidcast(
  source = "doctor-visits",
  signal = "smoothed_adj_cli",
  geo_type = "state",
  time_type = "day",
  geo_values = c("mi", "ny", "tx", "pa"),
  time_values = epirange(20200101, 20200601),
  as_of = as.Date("2020-06-01")
) |>
  as_epi_df()

bind_rows(
  edf_as_of |> mutate(version = "as_of arg"),
  snapshot_early|> mutate(version = "snapshot from epi_archive")
) |>
  filter(geo_value == "pa") |>
  ggplot(aes(x = time_value, y = value, 
  color = version, lty = version)) +
  geom_line() +
  labs(x = NULL, y = "Doctor visits (% CLI)",
    title = "pub_covidcast as_of vs. snapshot from epi_archive", 
    color = "",
    lty = "")
```

  - Use `revision_analysis()` to explore how signals change across revisions.

> **Hint:** use `$revision_behavior` on the object returned by `revision_analysis()` to access the summary statistics dataframe for each revision date (i.e. "key").

```{r}
#| label: revision-analysis
rev_result <- revision_analysis(archive_data, value)
rev_result$revision_behavior
```

> 🛈 A key metric generated by `revision_analysis()` is `lag_near_latest` (often called stabilization lag), which measures the number of days it takes for a data point to "settle" within a reliable threshold (e.g., 20%) of its final backfilled value.

> If the average stabilization lag is 20 days, what does that imply for a public health official trying to use yesterday's data to make a decision? How might you adjust your confidence in a signal based on these maturity metrics?

⚔️ Try out using the `within_latest` parameter within `revision_analysis()` to change the reliable threshold value. How does changing that value affect the output value of `lag_near_latest`?

```{r}
#| label: revision-within-latest
#| error: true
rev_tight <- revision_analysis(archive_data, value,
  within_latest = 0.1)

rev_loose <- revision_analysis(archive_data, value,
  within_latest = 0.3)

bind_rows(
  rev_tight$revision_behavior  |> mutate(threshold = "10%"),
  rev_result$revision_behavior |> mutate(threshold = "20%"),
  rev_loose$revision_behavior  |> mutate(threshold = "30%")
) |>
  group_by(threshold) |>
  summarize(median_lag = median(lag_near_latest, na.rm = TRUE))
```

⚔️ Use `dplyr`'s `group_by()` and `summarize()` to calculate revision statistics for each location.

```{r}
#| label: revision-by-state
rev_result$revision_behavior |>
  group_by(geo_value) |>
  summarize(
    mean_spread            = mean(spread, na.rm = TRUE),
    median_lag_near_latest = median(lag_near_latest, na.rm = TRUE)
    ) 
```
