---
title: "MINI-PROJECT 2: EDA and correlation analysis"
author: "Delphi Tooling Team"
date: today
execute:
  warning: false
  message: false
---

## Introduction

This notebook covers the second mini-project, focusing on exploratory data analysis (EDA) and correlation analysis of Delphi indicators.

## Load packages

`{InsightNetApr26}` package ensures all required Delphi tools and their correct versions/branches are installed.

```{r}
#| label: load-packages
if (!requireNamespace("pak", quietly = TRUE)) install.packages("pak")
if (!requireNamespace("InsightNetApr26", quietly = TRUE)) {
    pak::pkg_install("cmu-delphi/InsightNet-apr-2026/InsightNetApr26", dependencies = TRUE)
}
InsightNetApr26::verify_setup()

library(epidatr)
library(epiprocess)
library(dplyr)
library(ggplot2)
```

```{r}
#| label: setup
#| include: false
source("setup.R")
```

## Standard Timeframe

We define a standard timeframe for our analysis.

```{r}
#| label: timeframe
start_date <- 20220101
end_date <- 20220301
```

## General EDA utilities

First, we fetch a candidate indicator (Doctor visits CLI) to explore.

```{r}
#| label: fetch-candidate
res_cand <- pub_covidcast("doctor-visits", "smoothed_adj_cli", "state", "day",
    geo_values = c("mi", "ny", "tx", "pa"),
    time_values = epirange(start_date, end_date)
)
edf_cand <- as_epi_df(res_cand)

head(edf_cand)
```

### Sliding Averages and Growth Rates

We use `epi_slide` to calculate moving averages and growth rates.

```{r}
#| label: sliding
# Use epi_slide to calculate moving averages
edf_cand |>
    group_by(geo_value) |>
    epi_slide(smoothed_value = mean(value, na.rm = TRUE), .window_size = 7) |>
    autoplot(value, smoothed_value)

# Use epi_slide to calculate growth rate
edf_cand |>
    group_by(geo_value) |>
    mutate(growth_rate = growth_rate(value, time_value, method = "rel_change")) |>
    autoplot(value, growth_rate)
```

### Summary Statistics and Visualization

General summary statistics and data exploration using `autoplot` and `plot_heatmap`.

```{r}
#| label: eda-vis
# Get general summary statistics
summary(edf_cand)

# Use autoplot and plot_heatmap to explore the data.
autoplot(edf_cand, value, .facet_by = "geo_value")
autoplot(edf_cand, value, .interactive = T)
plot_heatmap(edf_cand, value)
```

## Compare two indicators

We fetch a second indicator (HHS COVID admissions) and join them into a single `epi_df`.

```{r}
#| label: compare-indicators
res_guid <- pub_covidcast("hhs", "confirmed_admissions_covid_1d", "state", "day",
    geo_values = c("mi", "ny", "tx", "pa"),
    time_values = epirange(start_date, end_date)
)
edf_guid <- as_epi_df(res_guid)

joined_edf <- edf_cand |>
    select(geo_value, time_value, candidate = value) |>
    full_join(
        edf_guid |> select(geo_value, time_value, guiding = value),
        by = c("geo_value", "time_value")
    ) |>
    as_epi_df()

# autoplot with two indicators
autoplot(joined_edf, candidate, guiding)
```

## Correlation Analysis

We calculate correlations using `epiprocess::epi_cor`.

```{r}
#| label: correlation
## Temporal correlation (within each location over time)
cor_temporal <- epi_cor(joined_edf, candidate, guiding, cor_by = geo_value)
head(cor_temporal)

## Spatial correlation (across all locations at each point in time)
cor_spatial <- epi_cor(joined_edf, candidate, guiding, cor_by = time_value)

ggplot(cor_spatial, aes(x = time_value, y = cor)) +
    geom_line() +
    labs(title = "Spatial correlation over time", y = "Correlation") +
    scale_y_continuous(limits = c(-1, 1))
```

### Lagged Correlation Analysis

We sweep lags from -14 to +14 days to see if the candidate leads the guiding indicator.

```{r}
#| label: lagged-correlation
# Sweep lags from -14 to +14 days
# Positive means candidate leads guiding
cor_lagged <- lapply(-14:14, \(x) {
    epi_cor(joined_edf, candidate, guiding,
        cor_by = geo_value, dt1 = -x
    ) |> mutate(lag = x)
}) |> bind_rows()

cor_lagged |>
    group_by(lag) |>
    summarize(mean_cor = mean(cor, na.rm = TRUE)) |>
    ggplot(aes(x = lag, y = mean_cor)) +
    geom_line() +
    geom_point() +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
    scale_y_continuous(limits = c(-1, 1)) +
    labs(title = "Mean correlation vs. lag", x = "Lag (days)", y = "Correlation")
```

## Revision behavior

Finally, we analyze the revision history of the candidate indicator.

```{r}
#| label: revision-analysis
# Fetch data versions by setting the issues parameter
archive_res <- pub_covidcast("doctor-visits", "smoothed_adj_cli",
    "state", "day",
    geo_values = c("mi", "ny", "tx", "pa"),
    time_values = epirange(start_date, end_date),
    issues = epirange(start_date, 20220401)
)

# For version processing
archive_data <- archive_res |>
    rename(version = issue) |>
    as_epi_archive()

# Plot the revision history
autoplot(archive_data, value, .versions = "week")

# Generate revision metrics
rev_metrics <- revision_analysis(archive_data, value)
rev_metrics
```
