---
title: "MINI-PROJECT 3: Customizing models"
author: "Delphi + MICOM"
date: today
execute:
  warning: false
  message: true
---

## Load packages

The `{InsightNetApr26}` package ensures all required Delphi tools and their correct versions/branches are installed.

```{r}
#| label: load-packages
if (!requireNamespace("pak", quietly = TRUE)) install.packages("pak")
if (!requireNamespace("InsightNetApr26", quietly = TRUE)) {
  pak::pkg_install("cmu-delphi/InsightNet-apr-2026/InsightNetApr26")
}
InsightNetApr26::verify_setup()

# If pak demands Rtools and you don't have it, you can use this instead:
#
# if (!requireNamespace("remotes", quietly = TRUE)) {
# install.packages("remotes")
# }
# remotes::install_github("cmu-delphi/InsightNet-apr-2026/InsightNetApr26")
# remotes::install_github("cmu-delphi/epidatr")
# remotes::install_github("cmu-delphi/epidatasets")
# remotes::install_github("cmu-delphi/epiprocess")
# remotes::install_github("cmu-delphi/epipredict")

library(epidatr)
library(epiprocess)
library(epipredict)
library(dplyr)
library(ggplot2)
library(parsnip)
```

```{r}
#| label: setup
#| include: false
source(here::here("scripts", "setup.R"))
```

## 3.1 Basic Modelling

⭐ We will use the built-in `covid_case_death_rates` dataset provided by the `epidatasets` package, which automatically loads with the other Delphi tools. 

> Hint: You can get a partial view of the dataset by entering `covid_case_death_rates` in the R console.

- Filter out the observations after August 1, 2021, to create a training set.

```{r}
#| label: training-set
# We define our forecast date and filter the data to only include
# what was available at that time.
forecast_date <- as.Date("2021-08-01")

training_set <- covid_case_death_rates |>
  # pretend we have one day of latency:
  filter(time_value <= forecast_date - 1L)
```

> ⓘ In a real-world scenario, reporting delays mean that recent data is often missing or subject to revision. If you accidentally trained your model on data that wasn't actually available at the time of the forecast, how would this "data leakage" give you a misleading impression of your model's accuracy?

::: {.callout-note}
This is why, for serious applications, we want to use `epi_archive`s and `epix_as_of` or `epix_slide` rather than the filtering approach above, in order to faithfully reflect the data that would have been fed into a model and how the model would have behaved.
:::

- Filter to four states of your choosing and use `autoplot()` to explore the `death_rate` and `case_rate` columns.

```{r}
#| label: exploration
# Pick four states and visualize the trends
target_states <- c("ca", "fl", "ny", "pa")

autoplot(training_set, death_rate, case_rate, .facet_filter = geo_value %in% target_states)
```


- Let’s fit an autoregressive model using lagged values of the `death_rate` column as features with `arx_forecaster()` on our training set.
> Hint: you need to provide the dataset and the outcome argument to `arx_forecaster()`

```{r}
#| label: basic-ar
# Fit a default AR model
fcst_ar <- arx_forecaster(
  training_set,
  outcome = "death_rate"
)

fcst_ar
```

- When you run a forecaster, it returns an object containing several useful components. 
  - Use `autoplot(your_forecast)` for a quick plot.
  - Use `your_forecast$predictions` on the object returned by the forecast to view the results. Use `pivot_quantiles_wider()` to expand the quantile distribution column into individual columns.
  - Use `your_forecast$epi_workflow` to access the underlying model. To see exactly how much weight the model gave to each lag, pass this object to `hardhat::extract_fit_engine()`.

```{r}
#| label: forecast-components
# Quick plot
autoplot(fcst_ar, .facet_filter = geo_value %in% target_states)

# Inspect predictions and pivot quantiles
fcst_ar$predictions |>
  pivot_quantiles_wider(.pred_distn) |>
  head()

# Access model weights
hardhat::extract_fit_engine(fcst_ar$epi_workflow)
```

> ⓘ How heavily does the model weigh data from exactly one week ago compared to yesterday? Try running `hardhat::extract_fit_engine(your_forecast$epi_workflow)` to examine the fitted coefficients for each lag. Which lag contributes the most to the forecast?

⚔️ Compare your autoregressive model fitted with `arx_forecaster()` to a simple model. The flatline forecaster is a common baseline that assumes the future will be exactly the last observed value, with uncertainty that grows as the forecast horizon grows. Fit a flatline model using `flatline_forecaster()` and examine the predictions to see how the quantiles behave.

```{r}
#| label: flatline-baseline
# Fit a flatline baseline model
fcst_flat <- flatline_forecaster(
  training_set,
  outcome = "death_rate"
)

fcst_flat

# Observe how quantiles widen over time
fcst_flat$predictions |>
  pivot_quantiles_wider(.pred_distn)
```

## 3.2 Model Customization

⭐ While “canned” forecasters such as `arx_forecaster()` are useful for establishing baselines, you will often want to customize your models. Use the `arx_forecaster()` arguments to customize your model. 

- So far, we have only used the outcome itself as a predictor. However, many signals serve as leading indicators. For example, changes in case rates often precede changes in death rates. Use both `case_rate` and `death_rate` as predictors of `death_rate` in `arx_forecaster()`. [Hint: use the `predictors` argument.]

```{r}
#| label: multi-predictor
# Use multiple signals to improve the forecast
fcst_multi <- arx_forecaster(
  training_set,
  outcome = "death_rate",
  predictors = c("case_rate", "death_rate")
)

fcst_multi

autoplot(fcst_multi, .facet_filter = geo_value %in% target_states)
```

- By default, `arx_forecaster()` uses linear regression. Swap this for `quantile_reg()`. 

```{r}
#| label: quantile-reg
# Use Quantile Regression for better uncertainty estimation
fcst_qr <- arx_forecaster(
  training_set,
  outcome = "death_rate",
  trainer = quantile_reg()
)

fcst_qr

autoplot(fcst_qr, .facet_filter = geo_value %in% target_states)
```

- Further customize your model by modifying the `arx_args_list()` arguments via the `args_list` parameter in `arx_forecaster()`. 

> Hint: you can access the `arx_args_list()` documentation by typing `?arx_args_list` in R.

  - Use the `lags` argument in `arx_args_list()` to specify provide a vector of integers representing the historical days for each indicator. 

> Hint: `lags = c(0, 7, 14)` tells the model to use data from today, one week ago, and two weeks ago as predictors. If you are using multiple indicators as predictors, you can provide a list of vectors (e.g., `lags = list(c(0, 7), c(0, 14))`) where each vector corresponds to one of your indicators.

  - Set the `ahead` argument to 28. 

```{r}
#| label: custom-args
# Customize lags and horizon
custom_args <- arx_args_list(
  lags = list(c(0, 7, 14), c(0, 7, 14)), # Lags for case_rate and death_rate
  ahead = 28
)

fcst_custom <- arx_forecaster(
  training_set,
  outcome = "death_rate",
  predictors = c("case_rate", "death_rate"),
  args_list = custom_args
)

fcst_custom

autoplot(fcst_custom, .facet_filter = geo_value %in% target_states)
```

⚔️ Create a forecast trajectory over a range of dates and plot it with `autoplot()`. This creates a full trajectory showing how uncertainty grows with the horizon.

> Hint: see this example on creating models for multiple ahead arguments: cmu-delphi.github.io/epipredict/articles/epipredict.html#generating-multiple-aheads

```{r}
#| label: trajectory-plot
# We loop over horizons to create a trajectory (0 to 28 days)
all_canned_results <- lapply(0:28, \(days_ahead) {
  arx_forecaster(
    training_set,
    outcome = "death_rate",
    predictors = c("case_rate", "death_rate"),
    trainer = quantile_reg(),
    args_list = arx_args_list(
      lags = list(c(0, 1, 2, 3, 7, 14), c(0, 7, 14)),
      ahead = days_ahead
    )
  )
})

# Extract and combine predictions from all horizons
results <- bind_rows(lapply(all_canned_results, \(x) x$predictions))

autoplot(
  object = all_canned_results[[1]]$epi_workflow,
  predictions = results,
  observed_response = covid_case_death_rates |>
    filter(time_value > "2021-07-01"), 
    .facet_filter = geo_value %in% target_states
)
```

## 3.3 Use Delphi data in other model types 

⭐ We will use `arx_classifier()` to predict disease growth categories. This model doesn't just predict a value, but it predicts the probability that the growth rate falls into specific categories.

- Create a classifier to predict whether `death_rate` is `<= 0.01` or `> 0.01` using `case_rate` and `death_rate` as predictors.

> Hint: to configure the classifier, we use `arx_class_args_list()` as input in the `args_list` argument. The key parameter is `breaks`, which defines the thresholds for our categories.

```{r}
#| label: classification-model
# Categorical prediction: is death_rate low (<= 0.01) or high (> 0.01)?

classifier_fcst <- arx_classifier(
  training_set |> filter(geo_value %in% c(tolower(state.abb), "dc")),
  outcome = "death_rate",
  predictors = c("case_rate", "death_rate"),
  args_list = arx_class_args_list(
    breaks = 0.01,
    ahead = 7
  )
)

classifier_fcst

# View the classification probabilities
classifier_fcst$predictions |>
  head()
```
