---
title: "MINI-PROJECT 1: Finding indicators and fetching data"
author: "Delphi Tooling Team"
date: today
execute:
  warning: false
  message: false
---

## Introduction

This notebook covers the first mini-project, focusing on finding indicators and fetching data through the Delphi Epidata API.

## Load packages

`{InsightNetApr26}` package ensures all required Delphi tools and their correct versions/branches are installed.

```{r}
#| label: load-packages
if (!requireNamespace("pak", quietly = TRUE)) install.packages("pak")
if (!requireNamespace("InsightNetApr26", quietly = TRUE)) {
    pak::pkg_install("cmu-delphi/InsightNet-apr-2026/InsightNetApr26")
}
InsightNetApr26::verify_setup()

library(epidatr)
library(epiprocess)
```

```{r}
#| label: setup
#| include: false
source("setup.R")
```

## Finding Indicators

You can explore available indicators in the [Delphi Epidata API documentation](https://cmu-delphi.github.io/delphi-epidata/api/covidcast-signals/doctor-visits.html). For this example, we'll use the "Doctor visits CLI" indicator.

## Fetching Data

We'll fetch the `smoothed_adj_cli` signal from the `doctor-visits` source for selected states over a specific time range.

```{r}
#| label: fetch-data
res <- pub_covidcast(
    source = "doctor-visits",
    signal = "smoothed_adj_cli",
    geo_type = "state",
    time_type = "day",
    geo_values = c("mi", "ny", "tx", "pa"),
    time_values = epirange(20220101, 20220301)
)
```

## Processing and Visualization

We convert the raw API response into an `epi_df` for easier analysis and plotting.

```{r}
#| label: process-data
# Convert the data into epi_df format
edf <- as_epi_df(res)

# View the header of the resulting epi_df
head(edf)

# Quickly plot the data using epiprocess::autoplot
autoplot(edf, value)
```
